---
title: "dissertation"
author: "Boiko Andrew"
date: "11 February 2021"
output:
  pdf_document:
    latex_engine: xelatex
  word_document: default
  html_document:
    df_print: paged
keep_tex: yes
keep_markdown: yes
header-includes:
- \XeTeXdefaultencoding utf8
- \usepackage{xltxtra}
- \usepackage{fontspec}
- \setmainfont{Times New Roman}
- \setsansfont{Arial}
- \setmonofont{Courier New}
- \newfontfamily{\cyrillicfont}{Times New Roman}
- \newfontfamily{\cyrillicfonttt}{Courier New}
- \newfontfamily{\cyrillicfontsf}{Arial}
- \usepackage[english]{babel}
---

```{r setup, eval = TRUE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

Remove all variables beforehand
```{r removing variables}
rm(list = ls())
```

Attach necessary packages
```{r attaching packages, echo = FALSE, message = FALSE}
library(rJava)
library(xlsx)
library(xts)
library(forecast)
library(dplyr)
library(rprojroot)
library(ggplot2)
```

Set up working directory
```{r determining directory, include = FALSE, cache = FALSE}
#knitr::opts_knit$set(root.dir =
#                       "D:/Мои проекты/Диссертация/R Studio") # дома
knitr::opts_knit$set(root.dir =
                       "H:/Разное/Мои проекты/Диссертация/R Studio/") # на работе
```

Read exchange rate from MS Excel file
```{r reading exchange rate, echo = TRUE, eval = TRUE}
fileName <- "RC_F01_01_2009_T09_02_2021.xlsx"
dailyCourse <- read.xlsx(file = fileName, sheetName = "RC",
                         encoding = "UTF-8")
```

```{r looking at the head of data, eval=TRUE}
head(as.matrix(dailyCourse))
```

Rename necessary columns and look at head data
```{r renaming columns, echo = TRUE}
colnames(dailyCourse)[colnames(dailyCourse) == "data"] <- "date"
colnames(dailyCourse)[colnames(dailyCourse) == "curs"] <- "rate"

dailyCourse$nominal <- NULL
dailyCourse$cdx <- NULL
```

Calculate daily сourse and plot it
```{r calculating daily course}
dailyCourseXts <- xts(x = dailyCourse$rate,
                      order.by = dailyCourse$date)
plot(dailyCourseXts, main = "Daily course")
```

Monthly course (mean value calculated over days)
```{r calculating monthly course}
monthlyCourseXts <- apply.monthly(dailyCourseXts, mean)
index(monthlyCourseXts) <- as.yearmon(index(monthlyCourseXts))
plot(monthlyCourseXts, main = "Monthly course")
```

Convert Xts object to time series object and plot it
```{r converting Xts object to time series object}
monthlyCourse <- ts(monthlyCourseXts, frequency = 12,
                    start = index(monthlyCourseXts)[1])
monthlyCourse <- monthlyCourse[ ,1]
plot(monthlyCourse, main = "Monthly course",
     xlab = "Years", ylab = "Course")
grid(col = "red")
```

Make some auxilary calculations
```{r auxilary calculations}
monthsQty <- as.integer(length(monthlyCourse))
yearsQty <- floor(length(monthlyCourse) / 12)
errorsQty <- 7 # ME, RMSE, MAE, MPE, MAPE, MASE, ACF1
startYear <-  as.integer(format(start(monthlyCourseXts), "%Y"))
```

# STL decomposition
Calculate number of differencings to convert time series into
stationary series in terms of seasonality
```{r number of seasonal differencings}
nsdiffs(monthlyCourse)
```

Calculate number of ordinal differencings to convert time series into
stationary series in term of mean and variance
```{r number of ordinal differencings}
ndiffs(monthlyCourse)
```

Plot the graph of time series over years
```{r seasonal plot}
seasonplot(monthlyCourse, 12, col = rainbow(12), year.labels = TRUE,
           main = "Seasonal plot")
grid(col = "black")
```

Seasonal decomposition itself
```{r seasonal decomposition}
monthlyCourseStl <- stl(monthlyCourse, s.window = "periodic")
par(mfrow = c(1,1))
plot(monthlyCourseStl)
grid(col = "red")
```

```{r trend plot}
plot(monthlyCourseStl$time.series[ , "trend"],
     xlab = "Years", ylab = "Stl trend component")
grid(col = "red")

```

```{r seasonal component plot}
plot(monthlyCourseStl$time.series[ , "seasonal"],
     xlab = "Years", ylab = "Stl seasonal component")
grid(col = "red")
```
``` {r remainder plot}
plot(monthlyCourseStl$time.series[ , "remainder"],
     xlab = "Years", ylab = "Stl remainder")
grid(col = "red")
```

# Arima
```{r}
windowWidth = 12  # window width is 12 months (1 year)

coefs <- c(0,0,0,0,0)
arimaForecast <- c(0)
for (i in 1:(monthsQty - windowWidth + 1)){
  subsetCourse <- subset(monthlyCourse, start = i,
                         end = (i + windowWidth - 1))
  varCoef <- sd(subsetCourse) / mean(subsetCourse)
  
  arimaModel <- auto.arima(subsetCourse)
  arimaCoefs <- arimaorder(arimaModel)

  arimaPred <- forecast(arimaModel)$mean[1]
  
  names(arimaCoefs) <- NULL
  arimaCoefs <- c(i, arimaCoefs, varCoef)

  if (i == 1){
    coefs <- arimaCoefs
    arimaForecast <- arimaPred
  }
  else{
    coefs <- rbind(coefs, arimaCoefs)
    arimaForecast <- rbind(arimaForecast, arimaPred)
  }
}

rownames(coefs) <- NULL
coefs <- as.data.frame(coefs)

names(coefs) <- c("n", "p", "d", "q", "varCoef")

coefs$pq <- coefs$p + coefs$q
coefs$sum <- coefs$p + coefs$d + coefs$q

varCoef <- coefs$varCoef
coefs$varCoef <- NULL
coefs$varCoef <- varCoef

rownames(arimaForecast) <- NULL
arimaForecast <- as.vector(arimaForecast)
```

```{r dataframe first rows}
head(as.matrix(coefs))
```

```{r d order plot}
plot(coefs$n, coefs$d, type = "h", main = "Arima d order plot",
     xlab = "Month", ylab = "d order")
grid(col = "red")
```
```{r p order plot}
plot(coefs$n, coefs$p, type = "h", main = "Arima p order plot",
     xlab = "Month", ylab = "p order")
grid(col = "red")
```
```{r q order plot}
plot(coefs$n, coefs$q, type = "h", main = "Arima q order plot",
     xlab = "Month", ylab = "p order")
grid(col = "red")
```

```{r p plus q order plot}
plot(coefs$n, coefs$pq, type = "h", main = "Arima p plus q order plot",
     xlab = "Month", ylab = "p plus q order")
grid(col = "red")
```

``` {r total order plot}
plot(coefs$n, coefs$sum, type = "h", main = "Arima total order plot",
     xlab = "Month", ylab = "Arima total order")
grid(col = "red")
```
```{r q vs. p plot (not jittered)}
plot(coefs$p, coefs$q, type = "p", main = "Arima q vs. p plot (not jittered)",
     xlab = "p order", ylab = "q order")
grid(col = "red")
```

```{r q vs. p plot (jittered)}
(p1 = ggplot(coefs, aes(x = p, y = q)) +
               geom_point(aes(fill = as.factor(d)), size = 3,
                          shape = 21, colour = "grey20",
             position = position_jitter(width = 0.2, height = 0.2)))
```
```{r count total rows of coefs dataframe}
as.matrix(coefs %>%
  summarise(count = n()))
```
```{r count different combinations of p, d and q}
as.matrix(coefs %>%
  group_by(p, d, q) %>%
  summarise(count = n()) %>%
  arrange(., desc(count)))
```

```{r plot monthly course along with coefficient of variation}
monthlyCourseCropped <- window(monthlyCourse,
                               start = c(startYear, windowWidth))
monthlyVarCoef <- ts(coefs$varCoef, start = c(startYear, windowWidth),
                     frequency = 12)

par(mfrow = c(2,1))

plot(monthlyCourseCropped, main = "Monthly course",
     xlab = "Years", ylab = "Course")
grid(col = "red")

plot(monthlyVarCoef, main = "Coefficient of variation",
     xlab = "Years", ylab = "q order")
grid(col = "red")
```
```{r plot monthly course along with coefficient of variation and sum of Arima coefficients}

monthlyCoefsSum <- ts(coefs$sum, start = c(startYear, windowWidth),
                      frequency = 12)

par(mfrow = c(3,1))

plot(monthlyCourseCropped, main = "Monthly course",
     xlab = "Years", ylab = "Course")
grid(col = "red")

plot(monthlyVarCoef, main = "Coefficient of variation",
     xlab = "Years", ylab = "q order")
grid(col = "red")

plot(monthlyCoefsSum, main = "Arima total order plot",
     xlab = "Years", ylab = "Arima total order")
grid(col = "red")
```

```{r plot the forecast itself}
plot(arimaForecast, type = "l", col = "red",
     main = "Arima forecast",
     xlab = "Years", ylab = "Exchange rate")
grid(col = "black")
```
```{r transform Arima forecast to timeseries object}
arimaForecast <- ts(arimaForecast, start = c(startYear, windowWidth),
                      frequency = 12)
```

```{r calculate accuracy}
accuracy(arimaForecast, monthlyCourseCropped)
```




